\texttt{JAXbind}: Bind any function to JAX

### 2024-06-27 paper/paper.tex

The use of JAX (Bradbury et al., 2018) is widespread in the natural
sciences. Of particular interest is JAX's powerful transformation
system. It enables a user to retrieve arbitrary derivatives of
functions, batch computations, and just-in-time compile code for
additional performance. Its transformation system requires that all
components of the computation are written in JAX.

@@

We believe \texttt{JAXbind} to be highly useful in scientific computing.
We intend to use this package to connect the Hartley transform and the
spherical harmonic transform from DUCC (Reinecke, 2024) to the
probabilistic programming package NIFTy (Edenhofer et al., 2024) as well
as the radio interferometry response from DUCC with the radio astronomy
package \texttt{resolve} (Arras et al., 2024). Furthermore, we intend to
connect the non-uniform FFT from DUCC with JAX for applications in
strong-lensing astrophysics. We envision many further applications
within and outside of astrophysics.

The functionality of \texttt{JAXbind} extends the external callback
functionality in JAX. Currently, \texttt{JAXbind}, akin to the external
callback functions in JAX, briefly requires Python's global interpreter
lock (GIL) to call the user-specified Python function. In contrast to
JAX's external callback functions, \texttt{JAXbind} allows for both a
custom Jacobian-vector product and vector-Jacobian product. To the best
of our knowledge no other code currently exists for easily binding
generic functions and both of their Jacobian products to JAX, without
the need for C++ or LLVM. The package that comes the closest is
Enzyme-JAX (W. S. Moses \& Zinenko, 2024), which allows one to bind
arbitrary LLVM/MLIR, including C++, with automatically-generated (W.
Moses \& Churavy, 2020; W. S. Moses et al., 2021, 2022) or
manually-defined derivatives to JAX.

PyTorch (Ansel et al., 2024) and TensorFlow (Abadi et al., 2015) also
provide interfaces for custom extensions. PyTorch has an extensively
documented Python interface\footnote{\url{https://pytorch.org/docs/stable/notes/extending.html}}
for wrapping custom Python functions as PyTorch functions. This
interface connects the custom function to PyTorch's automatic
differentiation engine, allowing for custom Jacobian and Jacobian
transposed applications, similar to what is possible with JAXbind.
Additionally, PyTorch allows a user to interface its C++ backend with
custom C++ or CUDA extensions\footnote{\url{https://pytorch.org/tutorials/advanced/cpp_extension.html}}.
JAXbind, in contrast, currently only supports functions executed on the
CPU, although the JAX built-in C++ interface also allows for custom GPU
kernels. TensorFlow includes a C++ interface\footnote{\url{https://www.tensorflow.org/guide/create_op}}
for custom functions that can be executed on the CPU or GPU. Custom
gradients can be added to these functions.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-tensorflow2015}{}%
Abadi, M., Agarwal, A., Barham, P., Brevdo, E., Chen, Z., Citro, C.,
Corrado, G. S., Davis, A., Dean, J., Devin, M., Ghemawat, S.,
Goodfellow, I., Harp, A., Irving, G., Isard, M., Jia, Y., Jozefowicz,
R., Kaiser, L., Kudlur, M., \ldots{} Zheng, X. (2015).
\emph{{TensorFlow}: Large-scale machine learning on heterogeneous
systems}. \url{https://www.tensorflow.org/}

\leavevmode\hypertarget{ref-PyTorch2024}{}%
Ansel, J., Yang, E., He, H., Gimelshein, N., Jain, A., Voznesensky, M.,
Bao, B., Bell, P., Berard, D., Burovski, E., Chauhan, G., Chourdia, A.,
Constable, W., Desmaison, A., DeVito, Z., Ellison, E., Feng, W., Gong,
J., Gschwind, M., \ldots{} Chintala, S. (2024, April). {PyTorch} 2:
Faster machine learning through dynamic {P}ython bytecode transformation
and graph compilation. \emph{29th ACM International Conference on
Architectural Support for Programming Languages and Operating Systems,
Volume 2 (ASPLOS '24)}. \url{https://doi.org/10.1145/3620665.3640366}

@@

\leavevmode\hypertarget{ref-Jax2018}{}%
Bradbury, J., Frostig, R., Hawkins, P., Johnson, M. J., Leary, C.,
Maclaurin, D., Necula, G., Paszke, A., VanderPlas, J., Wanderman-Milne,
S., \& Zhang, Q. (2018). \emph{{JAX}: Composable transformations of
{P}ython+{N}um{P}y programs} (Version 0.3.13) {[}Computer software{]}.
\url{http://github.com/google/jax}

@@

\leavevmode\hypertarget{ref-Moses2020}{}%
Moses, W., \& Churavy, V. (2020). Instead of rewriting foreign code for
machine learning, automatically synthesize fast gradients. In H.
Larochelle, M. Ranzato, R. Hadsell, M. F. Balcan, \& H. Lin (Eds.),
\emph{Advances in neural information processing systems} (Vol. 33, pp.
12472--12485). Curran Associates, Inc.
\url{https://proceedings.neurips.cc/paper/2020/file/9332c513ef44b682e9347822c2e457ac-Paper.pdf}

@@

\leavevmode\hypertarget{ref-Moses2024}{}%
Moses, W. S., \& Zinenko, O. (2024). \emph{{Enzyme-JAX}} (Version 0.0.6)
{[}Computer software{]}. \url{https://github.com/EnzymeAD/Enzyme-JAX}

\leavevmode\hypertarget{ref-ducc0}{}%
Reinecke, M. (2024). \emph{{DUCC}: Distinctly useful code collection}
(Version 0.33.0) {[}Computer software{]}.
\url{https://gitlab.mpcdf.mpg.de/mtr/ducc}

>>>>

The use of JAX (\protect\hyperlink{ref-Jax2018}{Bradbury et al., 2018})
is widespread in the natural sciences. Of particular interest is JAX's
powerful transformation system. It enables a user to retrieve arbitrary
derivatives of functions, batch computations, and just-in-time compile
code for additional performance. Its transformation system requires that
all components of the computation are written in JAX.

@@

We believe \texttt{JAXbind} to be highly useful in scientific computing.
We intend to use this package to connect the Hartley transform and the
spherical harmonic transform from DUCC
(\protect\hyperlink{ref-ducc0}{Reinecke, 2024}) to the probabilistic
programming package NIFTy
(\protect\hyperlink{ref-Edenhofer2023NIFTyRE}{Edenhofer et al., 2024})
as well as the radio interferometry response from DUCC with the radio
astronomy package \texttt{resolve}
(\protect\hyperlink{ref-Resolve2024}{Arras et al., 2024}). Furthermore,
we intend to connect the non-uniform FFT from DUCC with JAX for
applications in strong-lensing astrophysics. We envision many further
applications within and outside of astrophysics.

The functionality of \texttt{JAXbind} extends the external callback
functionality in JAX. Currently, \texttt{JAXbind}, akin to the external
callback functions in JAX, briefly requires Python's global interpreter
lock (GIL) to call the user-specified Python function. In contrast to
JAX's external callback functions, \texttt{JAXbind} allows for both a
custom Jacobian-vector product and vector-Jacobian product. To the best
of our knowledge no other code currently exists for easily binding
generic functions and both of their Jacobian products to JAX, without
the need for C++ or LLVM. The package that comes the closest is
Enzyme-JAX (\protect\hyperlink{ref-Moses2024}{W. S. Moses \& Zinenko,
2024}), which allows one to bind arbitrary LLVM/MLIR, including C++,
with automatically-generated (\protect\hyperlink{ref-Moses2021}{W. S.
Moses et al., 2021}, \protect\hyperlink{ref-Moses2022}{2022};
\protect\hyperlink{ref-Moses2020}{W. Moses \& Churavy, 2020}) or
manually-defined derivatives to JAX.

PyTorch (\protect\hyperlink{ref-PyTorch2024}{Ansel et al., 2024}) and
TensorFlow (\protect\hyperlink{ref-tensorflow2015}{Abadi et al., 2015})
also provide interfaces for custom extensions. PyTorch has an
extensively documented Python interface\footnote{\url{https://pytorch.org/docs/stable/notes/extending.html}}
for wrapping custom Python functions as PyTorch functions. This
interface connects the custom function to PyTorch's automatic
differentiation engine, allowing for custom Jacobian and Jacobian
transposed applications, similar to what is possible with JAXbind.
Additionally, PyTorch allows a user to interface its C++ backend with
custom C++ or CUDA extensions\footnote{\url{https://pytorch.org/tutorials/advanced/cpp_extension.html}}.
JAXbind, in contrast, currently only supports functions executed on the
CPU, although the JAX built-in C++ interface also allows for custom GPU
kernels. TensorFlow includes a C++ interface\footnote{\url{https://www.tensorflow.org/guide/create_op}}
for custom functions that can be executed on the CPU or GPU. Custom
gradients can be added to these functions.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-tensorflow2015}{}%
Abadi, M., Agarwal, A., Barham, P., Brevdo, E., Chen, Z., Citro, C.,
Corrado, G. S., Davis, A., Dean, J., Devin, M., Ghemawat, S.,
Goodfellow, I., Harp, A., Irving, G., Isard, M., Jia, Y., Jozefowicz,
R., Kaiser, L., Kudlur, M., … Zheng, X. (2015). \emph{{TensorFlow}:
Large-scale machine learning on heterogeneous systems}.
\url{https://www.tensorflow.org/}

\leavevmode\hypertarget{ref-PyTorch2024}{}%
Ansel, J., Yang, E., He, H., Gimelshein, N., Jain, A., Voznesensky, M.,
Bao, B., Bell, P., Berard, D., Burovski, E., Chauhan, G., Chourdia, A.,
Constable, W., Desmaison, A., DeVito, Z., Ellison, E., Feng, W., Gong,
J., Gschwind, M., … Chintala, S. (2024, April). {PyTorch} 2: Faster
machine learning through dynamic {P}ython bytecode transformation and
graph compilation. \emph{29th ACM International Conference on
Architectural Support for Programming Languages and Operating Systems,
Volume 2 (ASPLOS ’24)}. \url{https://doi.org/10.1145/3620665.3640366}

@@

\leavevmode\hypertarget{ref-Jax2018}{}%
Bradbury, J., Frostig, R., Hawkins, P., Johnson, M. J., Leary, C.,
Maclaurin, D., Necula, G., Paszke, A., VanderPlas, J., Wanderman-Milne,
S., \& Zhang, Q. (2018). \emph{{JAX}: Composable transformations of
{P}ython+{N}um{P}y programs} (Version 0.3.13).
\url{http://github.com/google/jax}

@@

@@

\leavevmode\hypertarget{ref-Moses2024}{}%
Moses, W. S., \& Zinenko, O. (2024). \emph{{Enzyme-JAX}} (Version
0.0.6). \url{https://github.com/EnzymeAD/Enzyme-JAX}

\leavevmode\hypertarget{ref-Moses2020}{}%
Moses, W., \& Churavy, V. (2020). Instead of rewriting foreign code for
machine learning, automatically synthesize fast gradients. In H.
Larochelle, M. Ranzato, R. Hadsell, M. F. Balcan, \& H. Lin (Eds.),
\emph{Advances in neural information processing systems} (Vol. 33, pp.
12472–12485). Curran Associates, Inc.
\url{https://proceedings.neurips.cc/paper/2020/file/9332c513ef44b682e9347822c2e457ac-Paper.pdf}

\leavevmode\hypertarget{ref-ducc0}{}%
Reinecke, M. (2024). \emph{{DUCC}: Distinctly useful code collection}
(Version 0.33.0). \url{https://gitlab.mpcdf.mpg.de/mtr/ducc}
